Conflict Driven Evolution and Urban Development: \\ Evidence from the Holy Roman Empire\footnote{ I thank Lara Andermann, Leonie Bielefeld, Daniel Bursian, Davide Cantoni, Emilio Esguerra, Filip Milojevic, Johannes Vorbach and Matthias Weigand for helpful comments and feedback. }

### 2023-06-27 paper/paper.tex
#31 almost finish conclusion

\begin{abstract}
How does switching between states affect the long-run development of cities, and why? I study this question in the setting of the Holy Roman Empire, using historical data on construction activity and the territorial history of cities. I specifically test the hypothesis that cities benefitted from ownership changes because competition between states selected for higher-quality government. While I show that cities grow faster in the long run after switching states, I do not find evidence that conflict-driven selection played a role in this.
\end{abstract}

@@

I build my paper on the theory of conflict driven evolution, outlined above, developed by \cite{levine2013, levine2021, levine2022}. Its main contribution is to explain why hegemonies arose in East Asia while balances of power could prevail in Europe: unlike China, the major powers of continental Europe were frequently preyed upon by strong outsiders --- the Vikings, the Swedes, and the English --- which prevented any one state from growing too strong (\citealp{levine2021}, p. 439). It is worth highlighting how the authors model war between states: you win them with state power, and state power is simply aggregate production in excess of the subsistence level. If the most successful states are the ones with the biggest surplus, selection should have made Europe increasingly prosperous.\footnote
{
The basic idea of Darwinian selection has previously been borrowed by economists - for example by \cite{galor2002} who argue that humans are subject to selection and accumulate (economically) favourable traits over the course of history. More infamously, \cite{clark2007} employs it in an attempt to explain why the Industrial Revolution took place in Europe, arguing that at some point the European rich began to out-breed the poor and passed their favourable traits on to their offspring.
}

@@

The closest relative of this paper is \cite{schoenholzer2022} who investigate how switching between states affects the development of cities. They find that switching states affects population negatively in the short run but positively in the long run. Further, they show that this long-run benefit is driven by improved state quality after switching. Similar to the evolution metaphor, they interpret this as "creative destruction": the short run cost of destruction is compensated by the long-run benefit of becoming part of a higher-quality state. However, they do not show explicitly how these improvements in state quality are driven by selection through conflict. They use a dataset with a bigger geographical scope (all of Europe) but lower resolution: they use urban population from \cite{bairoch1988} as a proxy for development, which is available for fewer cities and fewer years.

@@

While I am able to replicate \cite{schoenholzer2022}'s result that switching to another state has a positive effect one century later, I fail to identify a statistically significant effect of state power on urban development. This does not necessarily mean that conflict-driven evolution played no role in the rise of Europe --- there are fundamental issues with the data and the approach which limit the statistical power of my analysis. I argue that the effect would likely be significant in the absence of these issues.

@@

Early modern Europe was a uniquely warlike place. A recent theory models wars and conflict as a Darwinian selection mechanism: strong states absorb the territory of weak ones, which gradually filters out weaker states. If this is true, territory that changes hands during conflict will generally tend to move into the possession of a stronger state and benefit from its strength.

In this paper I test this prediction of the theory in the setting of the Holy Roman Empire. I want to understand the role that the process of conflict-driven evolution played in the economic development of Europe. I use data compiled from an encyclopedia of 2,390 historically German cities, covering the years 1300 to 1800. I use records of construction activity as a proxy for economic activity on the city level.

I run panel regressions with two-way fixed effects and staggered adoption on this dataset. First I estimate the dynamic effects of switching to another state. A significant effect manifests one century after a switch and becomes larger over time; after two centuries, we are 9 percentage points more likely to observe construction in cities that switched states.

I then estimate heterogeneous treatment effects of different types of switches. In particular, I care about the difference between the treatment effect of conquests and the treatment effect of successions: if we assume that conquests are more strongly associated with state power than successions, then we can interpret a switch by conquests as a proxy for an increase in state power.

What do we learn from this paper?

Accounting for the possibility of attenuation bias from measurement error makes the null result look less unexpected and more like suggestive evidence.

One could do more to explore the idiosyncrasies of the setting. 
Improvements / future research:
\begin{itemize}
    \item differences in timing - do early switchers in my sample experience different effects than late switchers?
    \item further explore differences across regions
    \item better measures of state power
\end{itemize}

>>>>

\begin{abstract}
How does switching between states affect the long-run development of cities, and why? I study this question in the setting of the Holy Roman Empire, using historical data on construction activity and the territorial history of cities. I specifically test the hypothesis that cities benefitted from ownership changes because competition between states selected for higher-quality government. While I show that cities grow faster in the long run after switching states, the evidence that conflict-driven selection played a role in this is less compelling.
\end{abstract}

@@

I build my paper on the theory of conflict driven evolution, outlined above, developed by \cite{levine2013, levine2021, levine2022}. Its main contribution is to explain why hegemonies arose in East Asia while balances of power could prevail in Europe: unlike China, the major powers of continental Europe were frequently preyed upon by strong outsiders --- the Vikings, the Swedes, and the English --- which prevented any one state from growing too strong (\citealp{levine2021}, p. 439). It is worth highlighting how the authors model war between states: you win them with state power, and state power is simply aggregate production in excess of the subsistence level. If the most successful states are the ones with the biggest surplus, selection should have made Europe increasingly prosperous.\footnote
{
The basic idea of Darwinian selection has previously been borrowed by economists - for example by \cite{galor2002} who argue that humans are subject to selection and accumulate (economically) favourable traits over the course of all history. More infamously, \cite{clark2007} employs it in an attempt to explain why the Industrial Revolution took place in Europe, arguing that at some point the European rich began to out-breed the poor and passed their middle-class values on to their offspring.
}

@@

The closest relative of this paper is \cite{schoenholzer2022} who investigate how switching between states affects the development of cities. They find that switching states affects population negatively in the short run but positively in the long run. Further, they show that this long-run benefit is driven by improved state quality after switching. Similar to the evolution metaphor, they interpret this as "creative destruction": the short run cost of destruction is compensated by the long-run benefit of becoming part of a higher-quality state. However, they do not show explicitly how these improvements in state quality are driven by selection through conflict. Their data has a bigger geographical scope (all of Europe) but lower resolution: they use urban population from \cite{bairoch1988} as a proxy for development, which is available for fewer cities and fewer years.

@@

While I am able to replicate \cite{schoenholzer2022}'s result that switching to another state has a positive effect one century later, I fail to identify a statistically significant effect of state power on urban development. This does not necessarily mean that conflict-driven evolution played no role in the rise of Europe --- there are fundamental issues with the data and the approach which limit the statistical power of my analysis. The most important one is attenuation bias from measurement error in the independent variable, and I argue that the effect may well be significant in its absence.

@@

Early modern Europe was a uniquely war-torn place. A recent theory models wars and conflict as a Darwinian selection mechanism: strong states absorb the territory of weak ones, which gradually filters out weaker states. If this is true, territory that changes hands during conflict will generally tend to move into the possession of a stronger state and subsequently benefit from its strength.

In this paper I test this prediction of the theory in the setting of the Holy Roman Empire. I want to understand the role that the process of conflict-driven evolution played in the economic development of Europe. I use data compiled from an encyclopedia of 2,390 historically German cities, covering the years 1300 to 1800 and featuring records of construction activity as a city-level proxy for economic activity. I run panel regressions with two-way fixed effects and staggered adoption on this dataset.

First I estimate the dynamic effects of switching to another state. A significant effect manifests one century after a switch and becomes larger over time; after two centuries, we are 9 percentage points more likely to observe construction in cities that switched states. I do not find a pre-trend.

I then estimate heterogeneous treatment effects of different types of switches. In particular, I care about the difference between the treatment effect of conquests and the treatment effect of successions. If we assume that conquered cities tend to move into the possession of stronger states than inherited cities, we can interpret the difference in treatment effects as the effect of state power. The literature suggests that state power is positively linked to growth-fostering institutions like fiscal capacity; we would therefore expect it to have a significant positive effect on economic (construction) activity.

I indeed obtain a positive estimate for the heterogeneity --- conquests increase the probability of observing construction activity by 4 percentage points more than successions --- but it is not statistically significant. This finding remains unchanged in most robustness checks. 
However, because the contrast between conquests and successions with respect to state power is not very sharp, all these estimates likely have attenuation bias working against them. Accounting for this possibility makes the null result seem less unexpected, and more like suggestive evidence: it means that the coefficient would be bigger and possibly significant if I had a more accurate measure of state power to regress on.

The main takeaway is that the effect of switching to another state on urban development is positive in the long run, and that I find tentative empirical support for the theory of conflict-driven selection.

It would be a big improvement on my research strategy to find a clearer measure of state power, to check whether measurement error really does bias my estimates downward. \cite{cantoni2023} construct a novel dataset of territory-level institutions which one could potentially use for that purpose. Another improvement would be to control for the number of cities switching to another state at the same time in Equation \eqref{eq:baseline}, as described in the Discussion. Third, an event study version of Equation \eqref{eq:baseline} may be able to better tease out significant effects if the effect of state power takes some time to phase in. Finally, recent literature (\cite{baker2022} provide an overview) has questioned the validity of standard two-way fixed effects regressions with staggered adoption and time-varying treatment --- one could improve my results by using one of the robust estimators proposed by \cite{sunab2021} or \cite{cengiz2019}, for example.

Future research in the setting of my thesis could explore the spatial heterogeneity for which I find evidence in Table \ref{tab:controls_50y} and Figure \ref{fig:conflict_map}. It would be interesting to learn why the coefficient on the conquest interaction is so dramatically different when the sample is skewed towards the southern and western regions. One could also take a much deeper dive into \cite{levine2021}'s theory and use the detailed geospatial and territorial histories data to test other implications of their model.

Better understanding how the European state system evolved can deepen our understanding of social and economic history and of the origins of efficient institutions.
